\bf Model-guided Concurrent Data Assimilation for Calibrating Cardiac Ion-channel Kinetics

### 2022-11-02 writing/draft.tex
Draft revision

	\if0\blind
	{
		\title{\bf Concurrent Data Assimilation for Modeling and Calibrating of Disease-altered Potassium Channel Kinetics in Mouse Cardiomyocytes}
		\author{Haedong Kim $^a$, Hui Yang $^a$, Andrew R. Ednie $^b$, and Eric S. Bennett $^b$ \\
		$^a$ Department of Industrial and Manufacturing Engineering, \\
		The Pennsylvania State University, University Park, PA USA \\
        $^b$ Department of Neuroscience, Cell Biology, and Physiology, \\
        Wright State University, Dayton, OH USA}
		\date{}
		\maketitle
	} \fi

@@

\begin{abstract}
Potassium channels (K\textsubscript{v}) are responsible for repolarizing the action potential in cardiomyocytes. Different K\textsubscript{v} isoforms and their currents contribute to various phases of repolarization. In the mouse ventricle, repolarization roughly consists of two kinetic components: the rapidly inactivating current I\textsubscript{Kto} (K\textsubscript{v}4.2), and the slowly inactivating currents I\textsubscript{Kslow1} and I\textsubscript{Kslow2} (K\textsubscript{v}1.5 and K\textsubscript{v}2.1). Data assimilation is used to estimate information on K\textsubscript{v} isoforms, because only the sum of their activities is measured in the form of currents (I\textsubscript{K}). Most existing studies separate K\textsuperscript{+} currents mathematically from I\textsubscript{K} recordings and make inference of kinetics by a curve-fitting decomposition procedure. However, this traditional approach has limitations to scrutinize the kinetics and dynamics of K\textsubscript{v} in two-fold: 1) the curve-fitting decomposition only relies on the shape of K\textsuperscript{+} currents without considering underlying kinetics and interactions of K\textsubscript{v} isoforms, and 2) it handles I\textsubscript{K} traces from the same cell one-by-one individually. There is an urgent need to develop new methods that simultaneously model the underlying kinetics of multiple I\textsubscript{K} traces for a better understanding of pathological mechanisms. We propose a novel \textit{concurrent data assimilation} framework that utilizes biophysics-based computer models of K\textsubscript{v} to delineate gating kinetics in the multiple observations concurrently to cope with cellular-level dynamics. We apply the proposed method to glyco-altered I\textsubscript{K} recordings from our previous studies. Experimental results show that the proposed method can discriminate characteristics between the control and experimental groups. This proposed method has strong potential to pave a new way to analyze K\textsubscript{v} and heart diseases.
\end{abstract}

\noindent%
{\it Keywords:} Data assimilation, computational electrophysiology, potassium channels, kinetics modeling, glycosylation

@@

\section{Introduction}
Voltage-gated potassium channels (K\textsubscript{v}) are critical to the electrical conduction system of the heart. K\textsubscript{v} are responsible for repolarizing the cardiac action potential (AP), a change of membrane potential over time, representing the net electrical activity in a myocyte. The AP is produced by the orchestrated activities of ion channels. Voltage-gated sodium channels (Na\textsubscript{v}) initiates fast depolarization that drives the sharp upstroke of the AP. K\textsubscript{v} play antagonistic roles that repolarize the AP with the rapid inactivation of Na\textsubscript{v}. Repolarization consists of multiple phases, to which different K\textsubscript{v} isoforms contribute with fluxes of Ca\textsuperscript{2+} and Cl\textsuperscript{-} involved through designated ion channels. This complex process can be affected by even modest changes in K\textsubscript{v} activities. Aberrant repolarization can significantly affect the AP duration and the QT interval, which lead to fatal heart diseases \citep{ravens2008role}. Glycosylation is a co/posttranslational modification that is critical for protein functions including activities of K\textsubscript{v} and other voltage-gated ion channels (VGICs) \citep{ohtsubo2006glycosylation,ednie2012modulation}. A growing number of studies in diverse directions, ranging from genome-wise to proteomic/glycomic searches, have shown clear links between altered protein glycosylation and heart diseases \citep{yung2004gene,yang2015glycoproteins,miura2016glycomics,nagai2016aberrant}. Despite its importance, the underlying mechanism is still largely unknown how changes in glycosylation contribute to heart disease onset and progression. Our \textit{in-vitro} experiments showed that glycosylation regulations modulate VGIC activities and contribute to both electrical and contractile dysfunction \citep{ednie2013sialicNav1,ednie2015sialicKv,ednie2019reduced}. In addition, we verified our experimental results via \textit{in-silico} studies and estimate how changes in the lower level (VGICs) affect the functions in the higher level (cardiomyocytes), which is one of the powers of systematic computer models \citep{du2013silico,du2015statistical,du2017silico,kim2022simulation}. \textbf{Figure~\ref{fig:framework_diagram}} illustrates this feedback loop of studying heart diseases, which will be referred to as the data assimilation framework that combines experimental data with mathematical/computational models.
\begin{figure}[!ht]
    \centering
    \includegraphics{figs/overall_diagram.pdf}
    \caption{Conceptual diagram to illustrate a data assimilation framework.}
    \label{fig:framework_diagram}
\end{figure}

However, a distinctive feature of K\textsubscript{v} raises challenges in both \textit{in-vitro} and \textit{in-silico} studies: K\textsubscript{v} have various isoforms playing different roles in AP repolarization \citep{nerbonne2005molecular}, whereas dominant isoforms exist for other major VGICs \citep{abriel2010cardiac,benitah2010type}. In the mouse ventricle, repolarization is roughly delineated as two kinetic components: the rapidly inactivating current I\textsubscript{Kto}, conducted through K\textsubscript{v}4.2, and the slowly inactivating currents such as I\textsubscript{Kslow1} and I\textsubscript{Kslow2}, conducted through K\textsubscript{v}1.5 and K\textsubscript{v}2.1 respectively. Although it is highly desirable to study individual K\textsubscript{v} isoforms to understand pathological mechanisms in the cardiac conduction system, only the collective activities as a whole can be measured in the form of currents using whole-cell voltage-clamp K\textsubscript{v} currents (I\textsubscript{K}) because of their slightly overlapping voltage-dependence of gating \citep{brouillette2004functional}. A data assimilation method, combining experimental data with mathematical/computational models, can used to estimate information on K\textsubscript{v} isoforms. Most existing studies separate individual K\textsuperscript{+} currents mathematically from I\textsubscript{K} by a curve-fitting decomposition procedure, assuming a curve of I\textsubscript{K} consists of the summation of exponential decay functions and a constant term; each represents the shape of K\textsuperscript{+} currents \citep{brunet2004heterogeneous}. However, this traditional approach is limited in its ability to scrutinize the kinetics and dynamics of K\textsubscript{v} in two-fold: 1) the curve-fitting I\textsubscript{K} decomposition only relies on the shape of K\textsuperscript{+} current without considering underlying kinetics and interactions of K\textsubscript{v} isoforms, and 2) it handles I\textsubscript{K} traces from the same cell with different conditions (i.e., clamp voltages) one-by-one individually. These limitations not only hinder \textit{in-silico} modeling results but, in some cases, make it impossible if modeling relies on kinetic data estimated from experimental data \citep{kim2022simulation}. There is an urgent need to develop new data assimilation methods that simultaneously delineate the underlying kinetics of multiple I\textsubscript{K} traces for a better understanding of the ion channels to provide insights into clinical applications, such as drug design.

Therefore, this paper proposes a new approach of \textit{concurrent data assimilation} for investigating the kinetics of K\textsubscript{v} isoforms from multiple I\textsubscript{K} traces measured in the same cell. A schematic flowchart of the proposed framework is shown in \textbf{Figure~\ref{fig:flow_chart}B} and compared with the current approach in \textbf{Figure~\ref{fig:flow_chart}A}. First, \textit{in-silico} models of K\textsubscript{v} and their currents are designed with parameters that control the kinetic rates and, in turn, determine the currents. It is worth mentioning that K\textsuperscript{+} currents are generated based on the mathematically simulated gating mechanism as opposed to the curve-fitting procedure, which only mimics the shape of the currents. Second, we perform a sensitivity analysis of the kinetic parameters using fractional factorial designs to identify the parameters that have significant impacts on the current generation. Several markers are defined that best represent the characteristics of the currents in various perspectives. Third, \textit{in-vitro} data of I\textsubscript{K} recordings are coupled with the proposed models by calibrating the parameters selected by the sensitivity analysis. We apply the proposed framework to glyco-altered IK recordings from our previous studies. Our data showed that reduced N-glycosylation in mouse cardiomyocytes is sufficient to cause dilated cardiomyopathy (DCM), which is the third most common cause of heart failure and the major reason for heart transplantation \citep{weintraub2017dilated}. Last, low-dimensional embedding is utilized to visually inspect the differences of the calibrated parameters in a collective way across the cells. Experimental results show that the proposed method provides reliable estimations of K\textsubscript{v} isoform kinetics of both the wild type and disease-altered groups and successfully captures cellular variability.
\begin{figure}
    \centering
    \includegraphics{figs/flow_chart.pdf}
    \caption{Flowcharts of (A) the current approach and (B) the proposed method.}
    \label{fig:flow_chart}
\end{figure}

The main advantage of this approach is that it analyzes multiple I\textsubscript{K} recordings concurrently based on the biophysical computer models, providing strong potential for modeling the kinetics and gating mechanism of the cardiac ion channels for studying heart diseases. Our contributions are summarized as follows:
\begin{itemize}
    \item We modeled the underlying kinetics of K\textsubscript{v} isoforms from whole-cell I\textsubscript{K} recordings based on biophysical principles rather than curve-fitting procedures.
    \item We developed a concurrent data assimilation routine that harnesses multiple protocols simultaneously rather than a single protocol independently one-by-one, resulting in comprehensive cellular-level modeling.
    \item The proposed approach generates separate K\textsubscript{v} models and visualize the distributions of the kinetic parameters for each cell, which allows analyzing the cellular variability. 
    \item Our method captures differences in channel kinetics given the disease-related perturbation, such as reduced glycosylation.
\end{itemize}
The proposed data assimilation framework has strong potential to serve as a novel complementary data analytical tool to \textit{in-vitro} experiments.

\section{Research Background}
\subsection{Computational Modeling of the Heart}
Computer models of the heart have been extensively studied and used in cardiology to understand the mechanism of the organ, investigate fatal diseases, and develop drugs. Models consist of mathematical equations describing biophysical properties and physiological functions compatible with experimental observations. Mathematical and computational implementations of the heart allow studying the heart in a quantitative and predictive way \citep{whittaker2020calibration}. For example, models can be used to interpolate processes not directly observed in experiments and extrapolate to novel conditions such as disease-related perturbations \citep{rodriguez2010systems}. With the advancement of sensing technology and data analysis, computational cardiology is expanding its field. Personalized heart models pave a new way for monitoring and managing the cardiac health of individuals by characterizing individual heart signals from the large-scale IoT network and calculating dissimilarities to detect potential cardiac risk \citep{yang2020network}. Hence, cardiac computer models provide a great opportunity to mitigate heart disease, the number one cause of death in the U.S. \citep{murphy2021mortality}.

The functionality of the heart to pump blood through the body is controlled by electrical signals generated in the heart itself. Hence, the electrical activities and waves spreading through the whole heart have been studied considerably. For examle, \citet{yang2016whole} a new reaction-diffusion and pattern recognition method to model spatiotemporal dynamics of electrical waves propagating throughout the heart surface. Sensing methods, such as vectorcardiogram (VCG) and electrocardiogram (ECG), measure the electrical activities of the heart. Various analytics methods have been proposed to cope with their complex nonlinear dynamics in different heart regions over time. \citet{chen2019heterogeneous} developed a heterogeneous recurrence method that characterizes spatiotemporal patterns in disease-altered multi-channel cardiac signals. Nonlinear recurrence analysis captures and quantifies topological relationships of recurring patterns in signals, which has been proven to be an effective tool for cardiac signals that are naturally recurring over time \citep{yang2010multiscale,meyers2020cross}.

However, to perform a fine-grained analysis, it is necessary to study the cellular-level activities of the heart as opposed to the organ-level signals. Voltage-gated ion channels (VGICs) are membrane proteins in cardiomyocytes that are the functional unit of the electrical activities of cardiac muscle cells. The whole-cell patch clamp recording is a powerful experimental technique for investigating the electrical properties of excitable cells such as cardiomyocytes and neurons. It allows measuring ionic currents conducted through designated VGICs and the action potential (AP). Although the laboratory technique reveals essential properties of VGICs and cardiomyocytes, the detailed gating mechanism remains ambiguous. For example, it is hard to determine rigorously how different ion channels collectively lead to changes in the AP. Computer models, coupled with experimental observations, provide integrative insights into the data, and they have become an essential part of studying the heart, as illustrated in Figure~\ref{fig:framework_diagram} \citep{winslow2011integrative}.

Hence, modeling VGICs involves a data assimilation procedure that combines experimental data with computer models. Traditionally, computer models are calibrated to characteristic curves estimated from ionic-current recordings \citep{colman2018description}. These estimation processes include curve fitting, which assumes a functional form of the characteristic curve and finds shape parameters best fit data. However, curve fitting only considers a single current recording at each time independently, so it cannot capture cellular-level dynamics from the current recordings by changes in input stimuli, i.e., voltage-clamp protocols. It raises an additional challenge for analyzing I\textsubscript{K} recordings, because one-more curve fitting is used to separate individual current traces generated by various K\textsubscript{v} isoforms. I\textsubscript{Ksum} traces are decomposed only based on the assumption of the shapes of the component currents but not the underlying biophysical gating kinetics \citep{kim2022simulation}. Therefore, there is an urgent need to develop a novel data assimilation method for modeling and calibrating K\textsubscript{v} kinetics with considering cellular dynamics.

\subsection{Altered Glycosylation and Heart Diseases}
Protein glycosylation is one of the most abundant and diverse forms of co/posttranslational modifications that impact essential protein functions, such as modulation of receptor or ion channel activities \citep{ohtsubo2006glycosylation,ednie2012modulation}. A growing number of studies has shown the association between altered glycosylation and heart diseases, such as dilated cardiomyopathy (DCM) and hypertrophic cardiomyopathy \citep{ohtsubo2006glycosylation,ednie2019reduced2}. It is reported that up to 20\% of patients with congenital disorders of glycosylation (CDG), who commonly have modest reductions in protein glycosylation, present with cardiac deficits and idiopathic DCM \citep{marques2017cardiac}. However, the underlying pathological mechanism still remains elusive.

We have investigated how regulated glycosylation contributes to heart failure in the context of electrophysiology. Electrical signaling is orchestrated activities of a variety of ion channels and transporters. Voltage-gated ion channels (VGICs) are heavily glycosylated, with ~30\% of the channel mass consisting of N-/O-linked glycans \citep{ednie2012modulation}. Glycosylation is a multi-step process and usually ends with sialic acid added. We also reported that a saturating, electrostatic effect of negatively charged sialic attached to the terminal of N-/O-glycan branches significantly altered electrical signaling in Na\textsubscript{v} \citep{ednie2013sialicNav1,ednie2015sialicNav2} as well as K\textsubscript{v} \citep{ednie2015sialicKv}. Computational modeling has been used to further investigate the functional role of reduced sialylation in Na\textsubscript{v} and K\textsubscript{v} \citep{du2015statistical, du2017silico}. However, these studies also rely on characteristic curves and statistics estimated from curve fitting.

\section{Methods}
\subsection{Computer Models of Potassium Channel Isoforms}
Mouse models have been very useful methods for studying human cardiac electric signaling \citep{nerbonne2004studying,milani2014small,kim2022irx5}. Computer models of mouse cardiomyocytes simulate important biophysical properties of ion channels and molecular dynamics in producing the AP. In this paper, the AP is modeled as a differential equation of transmembrane currents and stimulus current I\textsubscript{stim} as below in Equation~\ref{eq:ap} where $C_{m}$ is the membrane capacitance, $t$ is time.
\begin{equation}
    \label{eq:ap}
    \begin{split}
    -C_{m}\frac{dV}{dt} &= \mathrm{I}_{\mathrm{CaL}}+\mathrm{I}_{\mathrm{p}(Ca)}+\mathrm{I}_{\mathrm{NaCa}}+\mathrm{I}_{\mathrm{Cab}}+\mathrm{I}_{\mathrm{Na}}+\mathrm{I}_{\mathrm{Nab}}+\mathrm{I}_{\mathrm{NaK}}+\mathrm{I}_{\mathrm{Cl},Ca} \\
    &+\mathrm{I}_{\mathrm{Kto}}+\mathrm{I}_{\mathrm{Kslow1}}+\mathrm{I}_{\mathrm{Kslow2}}+\mathrm{I}_{\mathrm{Kss}}+\mathrm{I}_{\mathrm{Ks}}+\mathrm{I}_{\mathrm{Kr}}+\mathrm{I}_{\mathrm{K1}}+\mathrm{I}_{\mathrm{stim}}
    \end{split}
\end{equation}
There are 15 transmembrane currents in this AP model: the L-type calcium current (I\textsubscript{CaL}), the calcium pump current (I\textsubscript{p(Ca)}), the Na\textsuperscript{+}/Ca\textsuperscript{2+} exchange current (I\textsubscript{NaCa}), the calcium background current (I\textsubscript{Cab}), the fast Na\textsuperscript{+} current (I\textsubscript{Na}), the background Na\textsuperscript{+} current (I\textsubscript{Nab}), the Na\textsuperscript{+}/K\textsuperscript{+} pump current (I\textsubscript{NaK}), the Ca\textsuperscript{2+}-activated Cl\textsuperscript{-} current (I\textsubscript{Cl,Ca}), the rapidly inactivating transient outward K\textsuperscript{+} current (I\textsubscript{Kto}), the two ultra-rapidly activating delayed rectifier K\textsuperscript{+} currents (I\textsubscript{Kslow1} and I\textsubscript{Kslow2}), the non-inactivating steady-state K\textsuperscript{+} current (I\textsubscript{Kss}), the slow delayed rectifier K\textsuperscript{+} current (I\textsubscript{Ks}), the rapid delayed rectifier K\textsuperscript{+} current (I\textsubscript{Kr}), and the time independent K\textsuperscript{+} current (I\textsubscript{K1}). Detailed implementation of mouse AP and VGIC models can be found in \cite{bondarenko2014compartmentalized,asfaw2020compartmentalized}.

>>>>

	\if0\blind
	{
		\title{\bf Model-guided Concurrent Data Assimilation for Calibrating Cardiac Ion-channel Kinetics}
		\author{Haedong Kim $^a$, Hui Yang $^a$, Andrew R. Ednie $^b$, and Eric S. Bennett $^b$ \\
		$^a$ Department of Industrial and Manufacturing Engineering, \\
		The Pennsylvania State University, University Park, PA USA \\
        $^b$ Department of Neuroscience, Cell Biology, and Physiology, \\
        Wright State University, Dayton, OH USA}
		\date{}
		\maketitle
	} \fi

@@

\begin{abstract}
Potassium channels (K\textsubscript{v}) are responsible for repolarizing the action potential in cardiomyocytes. Different K\textsubscript{v} isoforms and their currents contribute to various phases of repolarization, e.g., the rapidly inactivating current I\textsubscript{Kto} (K\textsubscript{v}4.2), and the slowly inactivating currents I\textsubscript{Kslow1} and I\textsubscript{Kslow2} (K\textsubscript{v}1.5 and K\textsubscript{v}2.1). Because only the sum of their activities can be measured in the form of currents (I\textsubscript{Ksum}), there is a need to decompose and estimate different currents of K\textsubscript{v} isoforms. Most existing studies separate and make inference of K\textsubscript{v} kinetics via a curve-fitting procedure (e.g., tri-exponential fitting), but are limited in two-fold: 1) the curve-fitting decomposition only relies on the shape of K\textsuperscript{+} currents without considering underlying kinetics and interactions of K\textsubscript{v} isoforms; 2) I\textsubscript{Ksum} traces can only be decomposed and fitted for one clamp voltage at a time. In this paper, we develop a novel \textit{concurrent data assimilation} framework that calibrates biophysics-based computer models to delineate kinetic dynamics of various K\textsubscript{v} isoforms with multiple voltage-clamp responses simultaneously. The proposed method is evaluated and validated with voltage-clamp I\textsubscript{Ksum} recordings from our \textit{in-vitro} experiments on both wild-type and glyco-altered myocytes. Experimental results show that the proposed method effectively handles multiple responses in the voltage-clamp protocol and describes disease-altered perturbations to various K\textsubscript{v} isoforms. In addition, we develop a graphical-user-interface (GUI) software that provides an enabling tool to medical scientists for model calibration with voltage-clamp data. The proposed method and pertinent software are shown to have strong potential to study K\textsubscript{v} kinetics in variety of heart diseases.
\end{abstract}

\noindent%
{\it Keywords:} Data assimilation, simulation model, ion channels, congenital glycosylation disorder, model calibration

@@

\section{Introduction}
Cardiac voltage-gated potassium channels (K\textsubscript{v}) play critical roles in the electrical conduction system of the heart. K\textsubscript{v} are responsible for repolarizing the cardiac action potential (AP), a change of membrane potential over time representing the net electrical activity in a cardiomyocyte \citep{ravens2008role}. The shape and duration of the AP are primarily determined by various K\textsubscript{v} isoforms, contributing to different phases of repolarization with unique biophysical properties as illustrated in \textbf{Figure~\ref{fig:kv_ap}} \citep{nerbonne2005molecular}. Diverse distribution is a unique feature of K\textsubscript{v}. Dominant isoforms exist for other major voltage-gated ion channels (VGICs) such as Na\textsubscript{v} and Ca\textsubscript{v} \citep{abriel2010cardiac,benitah2010type}. Repolarization begins with the rapid inactivation of Na\textsubscript{v}, reducing inward Na\textsuperscript{+} flux, and fast open and close of K\textsubscript{v} isoforms, generating rapidly inactivating K\textsuperscript{+} current. There are other K\textsubscript{v} that remain open producing slowly inactivating currents, but in large species such as human, this phase 2 is characterized by prolonged repolarization due to intracellular Ca\textsuperscript{2+} flux (``plateau''). During phase 3, Ca\textsubscript{v} close and the cell begins to repolarize rapidly (``rapid repolarization''). In small species such as mice, phase 2 and 3 are not very distinctive. Hence, in the mouse ventricle, repolarization is roughly delineated as two kinetic components: the rapidly inactivating transient outward current I\textsubscript{Kto}, conducted through K\textsubscript{v}4.2, and the slowly inactivating delayed rectifier-type currents such as I\textsubscript{Kslow1} and I\textsubscript{Kslow2}, conducted through K\textsubscript{v}1.5 and K\textsubscript{v}2.1 respectively \citep{brunet2004heterogeneous}. Aberrant repolarization can significantly affect the AP duration and the QT interval, which lead to fatal heart diseases \citep{montpetit2009regulated}. 
\begin{figure}[!ht]
    \centering
    \includegraphics{figs/kv_ap.pdf}
    \caption{Ventricular action potential in mouse cardiomyocytes and the diverse K\textsubscript{v} isoforms and other ion channels contributing to shaping the AP.}
    \label{fig:kv_ap}
\end{figure}

However, only the collective activities of K\textsubscript{v} isoforms can be measured in the form of currents using whole-cell voltage-clamp recording (I\textsubscript{Ksum}) due to their slightly overlapping voltage dependence of gating \citep{brouillette2004functional}. Hence, there is a need to decompose I\textsubscript{Ksum} into individual K\textsuperscript{+} currents and estimate gating kinetics of K\textsubscript{v} isoforms to understand their pathological mechanisms in disease-altered cardiomyocytes. Data assimilation is a method of data analysis that combines experimental data with mathematical/computational models to make inferences on information of interest, such as individual K\textsuperscript{+} currents. Most existing studies separate individual K\textsuperscript{+} currents mathematically via a curve-fitting procedure, assuming a curve of I\textsubscript{Ksum} as the summation of exponential decay functions and a constant term; each represents the shape of individual K\textsuperscript{+} currents \citep{brunet2004heterogeneous}. However, this traditional approach is limited in its ability to scrutinize the kinetics and dynamics of K\textsubscript{v} in two-fold: 1) the curve-fitting decomposition is sorely determined by the shape of K\textsuperscript{+} currents without considering underlying kinetics and interactions of K\textsubscript{v} isoforms; 2) although a set of I\textsubscript{Ksum} traces are recorded with varying clamp voltages for a cell to investigate the voltage dependence of gating kinetics, only a single I\textsubscript{Ksum} trace is decomposed and fitted to individual K\textsuperscript{+} currents for one clamp voltage. ion methods that simultaneously delineate the underlying kinetics of multiple I\textsubscript{Ksum} traces for a better understanding of the ion channels to provide insights into mechanism of cardiomyocytes and clinical applications, such as drug design.

Therefore, this paper proposes a new approach of \textit{concurrent data assimilation} for investigating the kinetics of K\textsubscript{v} isoforms from multiple I\textsubscript{Ksum} traces measured in the same cell. First, \textit{in-silico} models of K\textsubscript{v} and their currents are designed with parameters that control the kinetic rates and, in turn, determine the currents. It is worth mentioning that K\textsuperscript{+} currents are generated based on the mathematically simulated gating mechanism as opposed to the curve-fitting procedure, which only mimics the shape of the currents. \textbf{Figure~\ref{fig:exp-vs-com}} shows two different approaches to assimilate the experimental data to extract information of K\textsubscript{v} isoforms. Second, we perform a sensitivity analysis of the kinetic parameters using fractional factorial designs to identify the parameters that have significant impacts on the current generation. Several markers are defined that best represent the characteristics of the K\textsuperscript{+} currents in various perspectives. Third, a calibration routine is proposed to adjust parameters selected by the sensitivity analysis to couple models with \textit{in-vitro} data of I\textsubscript{Ksum} recordings concurrently. Last, low-dimensional embedding is utilized to visually inspect the differences of the calibrated parameters in a collective way across the cells. We apply the proposed framework to glyco-altered I\textsubscript{Ksum} recordings from our previous studies. Glycosylation is a co/posttranslational modification that is critical for protein functions including activities of K\textsubscript{v} and other voltage-gated ion channels (VGICs) \citep{ohtsubo2006glycosylation,ednie2012modulation}. Our \textit{in-vitro} experiments showed that glycosylation regulations modulate VGIC activities and contribute to both electrical and contractile dysfunction \citep{ednie2013sialicNav1,ednie2015sialicKv,ednie2019reduced}. To be specific, reduced N-glycosylation in mouse cardiomyocytes is sufficient to cause dilated cardiomyopathy (DCM), which is the third most common cause of heart failure and the major reason for heart transplantation \citep{weintraub2017dilated}.
\begin{figure}[!ht]
    \centering
    \includegraphics{figs/exp-vs-com.pdf}
    \caption{Comparison between the traditional curve-fitting approach and the proposed model-guided data assimilation of \textit{in-vitro} experimental data of I\textsubscript{Ksum} recordings.}
    \label{fig:exp-vs-com}
\end{figure}

Experimental results show that the proposed method provides reliable estimations of K\textsubscript{v} isoform  currents and underlying kinetics of both the wild-type and disease-altered groups and successfully captures cellular variability. \textit{In-silico} modeling provides rich information that \textit{in-vitro} data cannot alone. For example, in our previous studies and this paper, computer models are used to verify experimental results, estimate channel kinetics, and show how changes in the lower level (VGICs) affect the functions in the higher level (cardiomyocytes), which is one of the powers of systematic computer models \citep{du2013silico,du2015statistical,du2017silico,kim2022simulation}. \textbf{Figure~\ref{fig:framework_diagram}} illustrates this feedback loop of studying heart diseases. Further, we develop a graphical-user-interface (GUI) application to make the proposed method accessible to medical scientists for investigating K\textsubscript{v} channelopathy, which is available as online supplemental material. The main advantage of the proposed approach is that it analyzes multiple I\textsubscript{Ksum} recordings concurrently based on the biophysical computer models, providing strong potential for modeling the kinetics and gating mechanism of the cardiac ion channels for studying heart diseases. Our contributions are summarized as follows:
\begin{itemize}
    \item We modeled the underlying kinetics of K\textsubscript{v} isoforms from whole-cell I\textsubscript{Ksum} recordings based on biophysical principles rather than curve-fitting procedures.
    \item We developed a concurrent data assimilation routine that harnesses a set of I\textsubscript{Ksum} traces simultaneously rather than a single current trace independently one-by-one, resulting in comprehensive cellular-level modeling.
    \item The proposed approach generates separate K\textsubscript{v} models and visualize the distributions of the kinetic parameters for each cell, which allows analyzing the cellular variability. 
    \item Our method captures differences in channel kinetics of the wild-type and the disease-altered groups.
\end{itemize}
\begin{figure}[!ht]
    \centering
    \includegraphics{figs/overall_diagram.pdf}
    \caption{Conceptual diagram to illustrate a data assimilation framework.}
    \label{fig:framework_diagram}
\end{figure}

\section{Research Background}
\subsection{Data Assimilation and Calibration of Cardiac Models}
The functionality of the heart to pump blood through the body is controlled by electrical signals generated in the heart itself. Cardiac electrophysiology has contributed significantly to our understanding of heart and disease-related modifications, such as congenital disorders of glycosylation (CDG). Advances in laboratory techniques allow studying molecular-level activities in cardiomyocytes via measuring currents of ion channels, which are functional units of the electrical conduction system of the heart. For example, whole-cell current recordings are used to show the roles of a certain gene, encoding $\alpha$ subunits that form Na\textsubscript{v} in human cardiomyocytes, which causes a small but inherent and chronic risk of acquired arrhythmia \citep{splawski2002variant}. Analysis of ionic currents showed pathophysiological roles of reduced sialylation impacting Na\textsubscript{v} and K\textsubscript{v} activities \citep{ednie2013sialicNav1, ednie2015sialicKv}; O-glycosylation modulations cause aberrant K\textsubscript{v} activities \citep{schwetz2011sialic}. However, \textit{in-vitro} experiments alone are limited to investigate channel activities. There is a need to assimilate experimental data to estimate information that cannot be observed directly. Traditionally, a curve-fitting method is used to make inferences, which assumes a functional form of information to estimate (e.g., individual currents of K\textsubscript{v} isoforms) \citep{liu2011dissection}.

However, the curve-fitting approach has limitations in its ability to assimilate the experimental data, because it only considers the shape of information to infer rather than reflecting underlying kinetics and interactions of various gating processes. Further, as curve fitting is only based on the shape of a single curve to fit, the traditional approach is not able to analyze a set of current recordings that are supposed to be analyzed together. In contrast, computer models of the heart consist of mathematical equations describing biophysical properties and physiological functions compatible with experimental observations. Mathematical and computational implementations of the heart allow studying the heart in a quantitative and predictive way \citep{whittaker2020calibration}. For example, models can be used to interpolate processes not directly observed in experiments and extrapolate to novel conditions such as disease-related perturbations \citep{rodriguez2010systems}. Computer models, coupled with experimental observations, provide integrative insights into the data by calibrating model parameters \citep{winslow2011integrative}. In the computational modeling literature, discrete-event simulation has been a primary focus, which only models events that occur apart over time. However, computer models of ion channels simulate gating kinetics that changes continuously, expressed mathematically in differential equations. The complex nonlinear structure of cardiac computer models also challenges calibration. However, little work has been done to calibrate continuous simulation-guided model.

\subsection{Congenital Disorders of Glycosylation}
Protein glycosylation is one of the most abundant and diverse forms of co/posttranslational modifications that impact essential protein functions, such as modulation of receptor or ion channel activities \citep{ohtsubo2006glycosylation,ednie2012modulation}. A growing number of studies has shown the association between altered glycosylation and heart diseases, such as dilated cardiomyopathy (DCM) and hypertrophic cardiomyopathy \citep{ohtsubo2006glycosylation,ednie2019reduced2}. It is reported that up to 20\% of patients with congenital disorders of glycosylation (CDG), who commonly have modest reductions in protein glycosylation, present with cardiac deficits and idiopathic DCM \citep{marques2017cardiac}. However, the underlying pathological mechanism still remains elusive.

We have investigated how regulated glycosylation contributes to heart failure in the context of electrophysiology. Electrical signaling is orchestrated activities of a variety of ion channels and transporters. Voltage-gated ion channels (VGICs) are heavily glycosylated, with ~30\% of the channel mass consisting of N-/O-linked glycans \citep{ednie2012modulation}. Glycosylation is a multi-step process and usually ends with sialic acid added. We also reported that a saturating, electrostatic effect of negatively charged sialic attached to the terminal of N-/O-glycan branches significantly altered electrical signaling in Na\textsubscript{v} \citep{ednie2013sialicNav1,ednie2015sialicNav2} as well as K\textsubscript{v} \citep{ednie2015sialicKv}. Computational modeling has been used to further investigate the functional role of reduced sialylation in Na\textsubscript{v} and K\textsubscript{v} \citep{du2015statistical, du2017silico}.

However, most of existing works utilized the curve-fitting method, e.g., bi-exponential or tri-exponential fitting, to assimilate the \textit{in-vitro} data from glycosylayion experiments. Curve-fitting provides estimates of characteristic curves (e.g., steady-state activation or inactivation) and statistics (e.g., time constants) for data-driven insights into CDG diseases. Few, if any, previous approaches have leveraged computer models to calibrate and delineate kinetic dynamics of various K\textsubscript{v} isoforms with multiple voltage-clamp responses simultaneously.

\section{Research Methodology}
\subsection{Computer Models of Potassium Channel Isoforms}
Mouse models have been very useful methods for studying human cardiac electric signaling \citep{nerbonne2004studying,milani2014small,kim2022irx5}. Computer models of mouse cardiomyocytes simulate important biophysical properties of ion channels and molecular dynamics in producing the AP. In this paper, the AP is modeled as a differential equation of transmembrane currents and stimulus current I\textsubscript{stim} as below in Equation~\ref{eq:ap} where $C_{m}$ is the membrane capacitance, $t$ is time.
\begin{equation}
    \label{eq:ap}
    \begin{split}
    -C_{m}\frac{dV}{dt} &= \mathrm{I}_{\mathrm{CaL}}+\mathrm{I}_{\mathrm{p}(Ca)}+\mathrm{I}_{\mathrm{NaCa}}+\mathrm{I}_{\mathrm{Cab}}+\mathrm{I}_{\mathrm{Na}}+\mathrm{I}_{\mathrm{Nab}}+\mathrm{I}_{\mathrm{NaK}}+\mathrm{I}_{\mathrm{Cl},Ca} \\
    &+\mathrm{I}_{\mathrm{Kto}}+\mathrm{I}_{\mathrm{Kslow1}}+\mathrm{I}_{\mathrm{Kslow2}}+\mathrm{I}_{\mathrm{Kss}}+\mathrm{I}_{\mathrm{Ks}}+\mathrm{I}_{\mathrm{Kr}}+\mathrm{I}_{\mathrm{K1}}+\mathrm{I}_{\mathrm{stim}}
    \end{split}
\end{equation}
There are 15 transmembrane currents in this AP model: the L-type calcium current (I\textsubscript{CaL}), the calcium pump current (I\textsubscript{p(Ca)}), the Na\textsuperscript{+}/Ca\textsuperscript{2+} exchange current (I\textsubscript{NaCa}), the calcium background current (I\textsubscript{Cab}), the fast Na\textsuperscript{+} current (I\textsubscript{Na}), the background Na\textsuperscript{+} current (I\textsubscript{Nab}), the Na\textsuperscript{+}/K\textsuperscript{+} pump current (I\textsubscript{NaK}), the Ca\textsuperscript{2+}-activated Cl\textsuperscript{-} current (I\textsubscript{Cl,Ca}), the rapidly inactivating transient outward K\textsuperscript{+} current (I\textsubscript{Kto}), the two ultra-rapidly activating delayed rectifier K\textsuperscript{+} currents (I\textsubscript{Kslow1} and I\textsubscript{Kslow2}), the non-inactivating steady-state K\textsuperscript{+} current (I\textsubscript{Kss}), the slow delayed rectifier K\textsuperscript{+} current (I\textsubscript{Ks}), the rapid delayed rectifier K\textsuperscript{+} current (I\textsubscript{Kr}), and the time independent K\textsuperscript{+} current (I\textsubscript{K1}). Detailed implementation of mouse AP and VGIC models can be found in \cite{bondarenko2014compartmentalized,asfaw2020compartmentalized}.
